---
title: "AME Example Solution"
author: "Prof. K."
date: '`r Sys.Date()`'
output: html_document
---

## Analysis

1. Open, save, and establish an R markdown document that includes a YAML header with your name, title, date and output as an html [1]. 

2. Include an R chunk that loads `tidyverse` [1].

```{r objectiv 2}
library(tidyverse)
```

3. Please download the data you see above by accessing [this link](data/bite_force.csv). Read it in as CSV data and save it to an object named `dat`. [3]

```{r objective 3}
dat <- read_csv("data/bite_force.csv")
```


4. Produce an illustrative visualization of the relationship between squirrel maximum bite force and mass and how this relationship may vary by sex. [3]

```{r objective 4}
dat %>% 
  ggplot(aes(mass,Fmax,col=sex))+geom_point()

```

## Interpretation

Overall, bite force increases with mass while males, on average have higher bite force than females. Perhaps the sexual differences arise due to sexual selection, i.e., males spar with one another in competition for females and over territory.


