---
title: "AME Example"
author: "A. Stew Dant"
date: '`r Sys.Date()`'
output: html_document
---

## Objectives

1. Open, save, and establish an R markdown document that includes a YAML header with your name, title, date and output as an html [1]. 

2. Include an R chunk that loads `tidyverse` [1].

```{r r objective 2, include=FALSE}
library(tidyverse)
```

3. Please download the data you see above by accessing [this link](data/bite_force.csv). Read it in as CSV data and save it to an object named `dat`. [3]

```{r objective 3}
dat <- read_csv("data/bite_force.csv")
```


4. Produce an illustrative visualization of the relationship between squirrel maximum bite force and mass and how this relationship may vary by sex. [3]

```{r}
dat_F <- dat %>% filter(sex=="Female")
dat_M <- dat %>% filter(sex=="Male")
dat_F %>% 
  ggplot(aes(mass,Fmax))+geom_point()
dat_M %>% 
  ggplot(aes(mass,Fmax))+geom_point()

```

## Interpretation

The relationship between bite force and mass is similar between males and females.


